Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2018-11-01 manuscript/Response_letter.Rmd

**You have our sincere thanks for your time and the very useful feedback, as well as these compliments.**

### Main comments

1) It is an issue common to all these "big data" studies, but the limitation should be better highlighted in the study. You are considering the lists of authors of articles in different journals, but you do not have information on individuals themselves. On of the main issues to identify the production of individuals is the presence of homonyms, and different ways to identify an author in a paper (e.g., F. Lastname vs Firstname Lastname).
Consequently, the results are about the correlation among genders of authors of papers, but not directly about individual preference.
For instance, you may have some individuals who are over-represented in your dataset (very prolific authors -- or powerful heads of institutions adding their names everywhere), and their presence on different papers/journals may be counted as independent units -- while they are not.
I am aware that the problem cannot be solved with ISI or Pubmed data, but it really needs to be acknowledged.

**We did not knowingly make any claims that pertain to individuals: we are well aware that because our study focuses on population-level patterns of assortment by gender, we can only speculate about the individual-level processes that produce the population-level patterns. Therefore, the fact that we cannot (or rather, did not try to) identify specific individuals does not compromise the paper's results and conclusions. We agree that some uses of the word 'preference' in our manuscript could be misleading in this respect, and so in the revision, we have replaced the word in key places (e.g. the title of the paper) with more neutral/agnostic terms (e.g. 'assortment').**

@@

- line 89 Author order differs across disciplines, e.g. it is alphabetical in maths. Please specify that you focus on life-sciences journals.

**We already specify that PubMed focuses on life sciences journals, as the reviewer notes elsewhere. Also, the presence of some alphabetically-ordered papers in the set would not alter the fact that the average seniority of last authors is greater than that of first authors, in most/all of the journals we studied.**

#### Methods
- line 265 Please specify how gender was assessed (even though this may already be in [5]). In particular, was it done automatically or manually? There may be issues with automatic gender assignment for authors of different geographic origins.

**We prefer to leave the full details to our previous paper, which has extensive descriptions, two different quality controls, and the full R code that was used to programmatically identify gender using authors' first names. Given our sample size (36 million authors), it hopefully goes without saying that automatic methods were used, but we now point this out explicitly in the Methods. Our earlier paper has extensive discussion of the geographic origin issue, and the proportion of authors to whom we could reliably assign a gender for each of >100 different countries (almost all countries had a gendering success rate of 70-90%, with China and Korea being the exceptions among the major research-producing nations).**

@@

- line 362 Why would senior authors be the main drivers? It could as well be junior authors who chose their advisor...

**Indeed, the purpose of the first author vs last author analysis was to see if we can get some clues as to whether homophily is driven mostly by senior or junior scientists: we outlined both possibilities in this section. I think you were misled because the two sentences beginning on line 362 use the word "Assume" rhetorically; the sentences basically say, "Assuming senior researchers drive the results, we predict X. This is because of Y." We do not start from the assumption that senior researchers are responsible for homophily. We also began that paragraph by pointing out that the 'senior researchers drive homophily' hypothesis is already predominant in the literature (though to our knowledge it has not been effectively tested before).**

- lines 397-407: Where do your data come from? actual proportions of women among early vs late career researchers?

**There is no data - it's a theoretical model, as explained in the section referenced here (e.g. the title says "Theoretical expectations"). We have re-written this section to make it even more obvious that there is switch from discussing data to discussing a model. For illustrative purposes, we investigated all combinations of values from 0% - 60% women, which spans >95% of journals on PubMed (see Holman et al. 2018).**

@@

#### Results
  - Figure 2:
  -- please use transparent shade so that the curve behind remains visible when it is lower than the curve in the front.

  **This is a stacked density plot, so there is no curve behind the other one (the blue is stacked on top of the white). We have clarified this in the figure legend.**

@@

-- Right-hand side figure: what do the dots correspond to? a single journal?

**Yes -- we now state this in the figure legend.**

  - Do you identify individual authors/ Control for the fact that some are more prolific than others?

**No, but this would not affect the veracity of our findings. Let's imagine that 90% of all papers in the world had one very prolific individual in their author lists. Let's assume this author is a man, though the argument would be the same if they were a woman. The existence of this prolific man would increase the proportion of male co-authors for both sexes, since most people have co-published with him. However, it would not affect our test statistic $\alpha'$, because the proportion of male coauthors would be elevated equally for men and women (and so $\alpha'$ would not change). You might ask, what if the super-author has preference for one gender of coauthors -- wouldn't that skew the results somehow? But then, $\alpha'$ would be working as intended: it would correctly record that homophilous (or heterophilous) co-authorships were more common than expected under random assortment by gender. This highlights a feature of $\alpha'$ noted earlier by the reviewer: it is a population-level measure, so we cannot make statements about individuals, e.g. we don't know if everyone is a little bit homophilous, or if there are a few prolific and very homophilous people (though the former is probably closest to the truth, since hyper-prolific authors with atypical levels of homophily are presumably the exception not the rule).**

### 2018-11-01 manuscript/Manuscript_Claire.Rmd
Ready for resubmission

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. Most journals had positive values of $\alpha'$ (77-92%, depending on time period and author type; S1 Data), and for many of these the false discovery rate (FDR)-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significant heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender. We also confirmed that in most journals (S2 Data) and most research disciplines (S3 Data, S1 Fig), the majority of papers had multiple authors. To verify the argument put forward in Figure 1, we also calculated $\alpha$ across all two-author papers in our entire 15-year PubMed dataset (excluding single-author papers, and papers where at least one author's gender was unknown; n = >3 million papers, >16 million authors). Across the whole dataset, $\alpha$ was 0.126, which is almost double the median value for individual journals (Figure 2; $\alpha = 0.070$ for 'All authors'), suggesting that lumping together papers from different fields and different time periods can indeed inflate $\alpha$.

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. In the stacked desnity plot, the white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (corrected p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. Points in the right panel show $\alpha'$ for individual journals.\label{alpha_histograms}](../figures/Fig2.pdf)

@@

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers. One possible explanation for this finding is that 2-authors papers might be more likely to have an author list that is evenly split between career stages (e.g. a postgraduate student and their supervisor), increasing the chance that the authors are mixed gender (see Figure 6). The result also suggests that the processes responsible for gender homophily are similar in small (e.g. 3 author) and larger (5+ authors) collaborations (and across disciplines where small versus large collaborations are the norm).

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
Given that we cannot identify individual researchers or their career stages, we used a simple model to derive the theoretical expectations for $\alpha$ when the gender ratio differs between career stages. As shown in \autoref{simulation}, we predict that $\alpha$ is expected to be non-zero, even if collaborators are randomly selected with respect to gender, provided that there is a gender gap between career stages. The extent to which $\alpha$ deviates from zero depends on the relative frequencies of collaboration within and between career stages (rows and columns in Figure 6), and the size of the gender gap between stages (x- and y-axes in Figure 6). When >50% of coauthor pairs comprise one early-career and one established researcher, we expect gender heterophily ($\alpha < 0$) whenever the gender ratio differs between career stages. Conversely, when >50% of collaborations are between people at the same career stage, we expect gender homophily ($\alpha > 0$). In a few parameter spaces (shown in red; \autoref{simulation}), $\alpha$ was quite high, and overlapped with the values that we estimated (\autoref{alpha_histograms}).

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. Contour lines denote increments of 0.1. \label{simulation}](../figures/Fig6_inkscape.pdf)

@@

Lastly, we note that if there is a gender gap between career stages and coauthorships between early-career and established researchers comprise >50% of the total, then the baseline expectation for $\alpha$ is actually less than zero (blue areas in \autoref{simulation}). Therefore, it is possible that researchers preferentially assort with same-gendered collaborators even more strongly than implied by other results, at least for certain journals or research disciplines.

@@

Regarding our finding of weaker homophily among 2-author papers, we suspect that many 2-author teams comprise a student/postdoc and a senior staff member, making these teams especially likely to be mixed-gender, due to the greater shortage of women among senior researchers [@Shaw_2012; @Holman_2018]. Assuming this interpretation is correct, this result suggests that our reported $\alpha'$ values may underestimate the strength of peoples' preferences for same-gendered collaborators; essentially, women seeking a senior collaborator could be constrained to work mostly with men, meaning that people's ideal and realised gender preferences would be mismatched. On a related note, Ghiasi et al. [-@Ghiasi_2015] argue that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women (for reference, their Figure 7 implies that coauthorships involving two women are 30% more frequent than expected under random assortment). By contrast, we feel that it may be counter-productive to recommend that women collaborate primarily with other women, e.g. because this constrains women's options (particularly in fields like engineering, where 90% of professors are men [@Shaw_2012]). Instead, we suggest that researchers of both genders can help to close the gender gap in STEMM. In the context of collaboration, one way to do this is to undertake self-examination to ensure that one is not inadvertently overlooking or excluding women among potential students and colleagues. One should also take care to treat male and female collaborators equally, e.g. in terms of training and mentoring, allocation of work, how one frames or promotes the collaboration (e.g. in conference presentations or on a website), and in terms of the credit afforded to collaborators' contributions. Experimental work suggests that unconscious bias causes people to undervalue women's research achievements [@Moss_2012], and a study of author contribution statements found observational evidence that menial or under-valued tasks are assigned to women while more prestigious tasks are assigned to men [@macaluso_2016].

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since the results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to gender differences in confidence, or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality (which is contentious; [@Garfield_2006]), our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In most STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers, due both to women leaving STEMM careers at greater rates (the 'leaky pipeline'), and to historical shortages of women studying STEMM ('demographic inertia') [@Shaw_2012; @Holman_2018]. We hypothesised that this difference in gender ratio between career stages could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations in a particular field are between students and professors, and that the gender ratio differs between career stages: we would then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no direct, causal effect. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

>>>>

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. Most journals had positive values of $\alpha'$ (77-92%, depending on time period and author type; S1 Data), and for many of these the false discovery rate (FDR)-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significant heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender. We also confirmed that in most journals (S2 Data) and most research disciplines (S3 Data, S1 Fig), the majority of papers had multiple authors.

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. In the stacked density plot, the white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (corrected p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. Points in the right panel show $\alpha'$ for individual journals.\label{alpha_histograms}](../figures/Fig2.pdf)

@@

Nevertheless, when we calculated $\alpha$ across all non-single-author papers in our entire 15-year PubMed dataset (as before, excluding papers where at least one author's gender was unknown; n = >3 million papers, >16 million authors), we found that $\alpha$ was 0.126. This figure is almost double the median value of $\alpha'$ for individual journals (Figure 2; $\alpha' = 0.070$ for 'All authors'), suggesting that lumping together papers from different fields and different time periods can indeed produce spurious evidence for gender homophily as outlined in Figure 1.

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplmentary Material). Papers with 3, 4 or $\ge5$ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers. One possible explanation for this finding is that 2-authors papers might be more likely to have an author list that is evenly split between career stages (e.g. a postgraduate student and their supervisor), increasing the chance that the authors are mixed gender (see Figure 6). The result also suggests that the processes responsible for gender homophily are similar in small (e.g. 3 author) and larger (5+ authors) collaborations (and across disciplines where small versus large collaborations are the norm).

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
Given that we cannot identify individual researchers or their career stages, we used a simple model to derive the theoretical expectations for $\alpha$ when the gender ratio differs between career stages (see Methods). As shown in \autoref{simulation}, we predict that $\alpha$ is expected to be non-zero, even if collaborators are randomly selected with respect to gender, provided that there is a gender gap between career stages. The extent to which $\alpha$ deviates from zero depends on the relative frequencies of collaboration within and between career stages (rows and columns in Figure 6), and the size of the gender gap between stages (x- and y-axes in Figure 6). When >50% of coauthor pairs comprise one early-career and one established researcher, we expect gender heterophily ($\alpha < 0$) whenever the gender ratio differs between career stages. Conversely, when >50% of collaborations are between people at the same career stage, we expect gender homophily ($\alpha > 0$). In a few parameter spaces (shown in red; \autoref{simulation}), $\alpha$ was quite high, and overlapped with the values that we estimated (\autoref{alpha_histograms}).

![When the gender ratio of early-career researchers is not equal to the gender ratio among established researchers, the null expectation for $\alpha$ is not necessarily zero. Specifically, if most collaborations occur between career stages, there will be an excess of mixed-gender collaborations ($\alpha < 0$, blue areas), while if most collaborator pairs comprise two people at the same career stage, there will be an excess of same-gender collaborations ($\alpha > 0$, red areas). However, the conditions required for strong gender homophily (i.e. the red areas) are quite restrictive, making it unlikely that this issue can fully explain the homophily observed in our study. Additionally, in research disciplines where between-career stage collaboration is common and there is a shortage of women among established researchers (i.e. the blue areas), our study will underestimate the strength of gender homophily. Contour lines denote increments of 0.1. \label{simulation}](../figures/Fig6_inkscape.pdf)

@@

Lastly, we note that if there is a gender gap between career stages and coauthorships between early-career and established researchers comprise >50% of the total, then the baseline expectation for $\alpha$ is actually less than zero (blue areas in \autoref{simulation}). Therefore, it is possible that researchers preferentially assort with same-gendered collaborators even more strongly than implied by our results, at least for certain journals or research disciplines.

@@

Regarding our finding of weaker homophily among 2-author papers, we suspect that many 2-author teams comprise a student/postdoc and a senior staff member, making these teams especially likely to be mixed-gender, due to the greater shortage of women among senior researchers [@Shaw_2012; @Holman_2018]. Assuming this interpretation is correct, this result suggests that our reported $\alpha'$ values may underestimate the strength of peoples' preferences for same-gendered collaborators; essentially, women seeking a senior collaborator could be constrained to work mostly with men, meaning that people's ideal and realised gender preferences would be mismatched. On a related note, Ghiasi et al. [-@Ghiasi_2015] argue that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women (for reference, Figure 7 in [-@Ghiasi_2015] implies that coauthorships involving two women are $c$. 30% more frequent than expected under random assortment). By contrast, we feel that it may be counter-productive to recommend that women collaborate primarily with other women, e.g. because this constrains women's options (particularly in fields like engineering, where 90% of professors are men [@Shaw_2012]). Instead, we suggest that researchers of both genders can help to close the gender gap in STEMM. In the context of collaboration, one way to do this is to undertake self-examination to ensure that one is not inadvertently overlooking or excluding women among potential students and colleagues. One should also take care to treat male and female collaborators equally, e.g. in terms of training and mentoring, allocation of work, how one frames or promotes the collaboration (e.g. in conference presentations or on a website), and in terms of the credit afforded to collaborators' contributions. Experimental work suggests that unconscious bias causes people to undervalue women's research achievements [@Moss_2012], and a study of author contribution statements found observational evidence that menial or under-valued tasks are assigned to women while more prestigious tasks are assigned to men [@macaluso_2016].

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? We beleive that the answers to these questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since the results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to gender differences in confidence, or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality (which is contentious; [@Garfield_2006]), our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In most STEMM disciplines, the gender ratio is more skewed among established researchers relative to early-career researchers, due both to women leaving STEMM careers at greater rates (the 'leaky pipeline'), and to historical shortages of women studying STEMM subjects at university ('demographic inertia') [@Shaw_2012; @Holman_2018]. We hypothesised that this difference in gender ratio between career stages could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations in a particular field are between students and professors, and that the gender ratio differs between career stages: we would then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no direct, causal effect. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).
